---
title: "Reproducible Research: Peer Assessment 1"
Author: "Matthew Renze"
Date: "02/14/2015"
output: 
  html_document:
    keep_md: true
---
Author: Matthew Renze  
Date: 02/14/2015

## Loading and preprocessing the data

```{r load}
# Load the data
data <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?
```{r totals, message=FALSE}
# Load dplyr library
library(dplyr)

# Get sum of steps by day
totals <- data %>%
  do(filter(., complete.cases(.))) %>%
  group_by(date) %>% 
  summarise(steps = sum(steps))

# Create a histogram of the total number of steps per day
hist(
  x = totals$steps,
  main = "Histogram of Total Number of Steps per Day",
  xlab = "Number of Steps per Day")

# Calculate the mean and median steps per day
meanSteps <- mean(totals$steps)
medianSteps <- median(totals$steps)
```
The mean number of steps per day is `r format(meanSteps, big.mark=",", nsmall = 2)`.  
The median number of steps per day is `r format(medianSteps, big.mark=",")`.


## What is the average daily activity pattern?
```{r averages, message=FALSE, warning=FALSE}
# Get average steps by interval
averages <- data %>% 
  do(filter(., complete.cases(.))) %>%
  group_by(interval) %>% 
  summarise(steps = mean(steps))

# Plot time series of average steps by interval of day
plot(
  x = averages$interval,
  y = averages$steps,
  type = "line",
  main = "Average Daily Steps by Interval of Day",
  xlab = "5-minute Interval of Day",
  ylab = "Average Number of Steps")

# Get the interval with the max average steps
max <- averages %>% 
  filter(steps == max(steps))

```
The 5-minute interval with the maximum number of steps across all days is interval `r max$interval` with `r format(max$steps, digits = 5, nsmall = 2)` steps.

## Imputing missing values
For our strategy to impute missing values, we will use the average value across all days for the corresponding missing 5-minute interval.
```{r na, message=FALSE}
# Count NA values
nas <- sum(is.na(data$steps))
```
There are `r nas` rows containing missing (NA) values.

```{r impute, message=FALSE}
# Copy raw data
imputed <- data.frame(data)

# Impute missing values using average daily steps
for(i in 1:nrow(imputed)){
    if(is.na(imputed[i, "steps"])){
      imputed[i, "steps"] <- data %>% 
        do(filter(., complete.cases(.))) %>%
        filter(interval == imputed[i, "interval"]) %>%
        group_by(interval) %>% 
        summarise(steps = mean(steps)) %>%
        select(steps)
  }
}

# Get sum of steps by day (imputed)
imputedTotals <- imputed %>%
  do(filter(., complete.cases(.))) %>%
  group_by(date) %>% 
  summarise(steps = sum(steps))

# Create a histogram of the total number of steps per day (imputed)
hist(
  x = imputedTotals$steps,
  main = "Histogram of Total Number of Steps per Day (Imputed)",
  xlab = "Steps per Day (Imputed)")

# Calculate the mean and median steps per day (imputed)
imputedMeanSteps <- mean(imputedTotals$steps)
imputedMedianSteps <- median(imputedTotals$steps)
```
The mean number of steps per day (imputed) is `r format(imputedMeanSteps, big.mark=",", nsmall = 2)`.  
The median number of steps per day (imputed) is `r format(imputedMedianSteps, big.mark=",")`.  

So, when using imputed values, the mean stayed the same and the median was moved to a day containing all imputed values.  

## Are there differences in activity patterns between weekdays and weekends?

```{r days, message=FALSE}
# Define weekends
weekends = c("Saturday", "Sunday")

# Add a factor variable for type of day
for(i in 1:nrow(imputed)){
  if (weekdays(as.Date(imputed[i, "date"])) %in% weekends) {
    imputed[i, "type"] = "Weekend"
  }
  else {
    imputed[i, "type"] = "Weekday"
  }
}

# Convert day type to factor
imputed$type <- as.factor(imputed$type)

# Get average steps by interval
imputedAverages <- imputed %>% 
  group_by(interval, type) %>% 
  summarise(steps = mean(steps)) %>%
  arrange(type, interval)

# Load lattice library
library(lattice)

# Create plot
xyplot(
  steps~interval|type,
  data = imputedAverages,
  type = "l",
  main = "Average Steps by 5-Minute Interval",
  xlab = "5-minute Interval",
  ylab = "Number of Steps (imputed)",
  layout=c(1,2))
```
